Clustering Beat-Chroma Patterns\\ in a Large Music Database

### 2010-03-25 Paper/Paper.tex
figure figure 1 placement
changed all figure placement to [t]
more rewording to sections 5 and 6

\begin{figure}[thb]
\begin{center}
\includegraphics[width=.4\columnwidth]{code}
\end{center}
\caption{\small{A typical codeword from a codebook of size $200$ (code
    7 in Figure \ref{fig:codes1}), corresponding to a
    tonic-subdominant chord progression.  The patch is composed of $2$
    bars and the pattern length was set to $8$ beats.  }}
\label{fig:code}
\end{figure}

@@

\begin{figure}[b]
\begin{center}
\includegraphics[width=.8\columnwidth]{encoding_per_beat}
\end{center}
\caption{\small{Encoding patterns of different sizes with a fixed size
codebook of 100 patterns. The size of the pattern is defined by the number of beats. 
Downbeat (bar alignment) information was not used for this experiment. % why not?
}}
\label{fig:perbeat}
\end{figure}

@@

\begin{figure}[bt]
\begin{center}
\includegraphics[width=.99\columnwidth]{freqs}
\end{center}
\caption{\small{Usage proportions for all 200 codewords 
on the {\it artist20} test set (which comprises $71,832$
patterns).  Also shown are the usage proportions for the training set (``cowbell''), 
which are similar.
Note that even though all codewords are initialized from samples, some
are used only once in the training set, or not at all for test set.
}}
\label{fig:freqs}
\end{figure}

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.9\columnwidth]{codes_lle}
\end{center}
\caption{\small{ Locally linear embedding (LLE) visualization of the
    codebook described in Section \ref{sec:visu}.  Each codeword is
    represented by a point on the scatter plot.  Randomly selected
    codewords from the embedded space are displayed over the plot.
    %FIXME: does this figure really provide any new insight?
  }}
\label{fig:lle}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.9\columnwidth]{code_variance}
\end{center}
\caption{\small{Average variance in time of a codebook. The vertical
axis cuts at the $63$rd pattern, a rough estimate of the number of 
codewords consisting of sustained chords.
}}
\label{fig:code_var}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.8\columnwidth]{codes_8bars}
\end{center}
\caption{\small{Sample of longer codewords spanning over $8$
    bars. Codewords selected randomly from the codewords used in the
    test set. Percentage of use display in parentheses.  Most patterns
    consist of sustained note or chords, but code 0 shows a repeating
    two chord pattern and code 4 contains two bars of a
    1$\rightarrow$1$\rightarrow$1$\rightarrow$2 progression.}}
\label{fig:codes8bars}
\end{figure}

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.9\columnwidth]{close_patterns1}
\end{center}
\caption{\small{Cluster around centroid presented in
Figure \ref{fig:code}. Taken from the artist20 dataset, the cluster
size is actually $639$. Shown samples were randomly selected.
}}
\label{fig:cluster}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.8\columnwidth]{close_patterns_diff}
\end{center}
\caption{\small{First three patterns of figure \ref{fig:cluster}
($2$nd line) presented with the centroid from Figure \ref{fig:code}
($1$st line) and the absolute difference between both ($3$rd line).
}}
\label{fig:cluster_diff}
\end{figure}

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.9\columnwidth]{song_encoded}
\end{center}
\caption{\small{Good Day Sunshine by \textit{The Beatles}.
Original song and its encoding using a $200$ entry codebook learning
$2$ bars at a time.
}}
\label{fig:encodesong}
\end{figure}

\section{Applications}\label{sec:exps2}
In this section we apply the result of our method to two tasks
unrelated to clustering. We try to demonstrate that the codewords
we learned capture musically meaningful patterns, at least to
contemporary pop music. This also demonstrates the utility of unsupervised 
clustering to other MIR tasks usually tackled in a supervised
framework. In the first experiment, we show that bar information is retained 
in the codewords. Thus we can discover the bar boundaries for new songs.
The second experiment shows we can discriminate artists
based on the encoding of their songs, hence we can associate artists
with specific codes.

@@

\begin{table}
\begin{center}
\begin{tabular}{cc}
\toprule
Offset & \% of times chosen \\
\midrule
0 & $\mathbf{62.6}$\\
1 & $16.5$\\
2 & $09.4$\\
3 & $11.5$\\
\bottomrule
\end{tabular}
\end{center}
\caption{\small{
Results from the song segmentation experiment.
The correct offset is $0$.
Baseline performance by choosing the offset randomly gives an accuracy of $25$\%.
}}
\label{tab:offset}
\end{table}

@@

\begin{figure}[htb]
%\begin{center}
\includegraphics[width=.9\columnwidth]{conf_mat_per_artist}
%\end{center}
\caption{\small{Confusion matrix for artist recognition task.}}
\label{fig:conf_mat}
\end{figure}

@@

% ignoring the two following figures
\iffalse
\begin{figure}[htb]
\begin{center}
\includegraphics[width=.4\columnwidth]{metallica_pattern}
\end{center}
\caption{\small{Most typical pattern for Metallica.
COULD BE MIXED WITH TORI AMOS FIGURE.
}}
\label{fig:metallica}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.4\columnwidth]{toriamos_suzannevega_pattern}
\end{center}
\caption{\small{Most typical pattern for Tori Amos and Suzanne Vega.
    Can we show the top 2 or 4 for each to get a better idea?}}
\label{fig:amosvega}
\end{figure}
\fi

\begin{figure}[htb]
  \centering
  \subfloat[][\small{Metallica}]{\label{fig:metallica2}\includegraphics[width=0.32\columnwidth]{metallica_pattern}}
  \hspace{5mm}                
  \subfloat[][\small{Tori Amos / Suzanne Vega}]{\label{fig:amosvega2}\includegraphics[width=0.32\columnwidth]{toriamos_suzannevega_pattern}}
  \vspace{2mm}
  \caption{\small{Typical patterns for different artists.}}
  \label{fig:typicalpat}
\end{figure}

\section{Conclusion and Future Work}
We presented a practical method to perform large-scale clustering of
harmonic patterns. We assessed the basic properties of the method through
experiments on a large collection of music. We suggested many ways
to interpret the data (such as LLE) and assert the merit of the codebook
through further experiments.
We discussed the possibility to move to even larger scales
and we provided our source code.\footnote{Code not yet released to preserve
submission's anonymity}

>>>>

\begin{figure}[t]
\begin{center}
\includegraphics[width=.4\columnwidth]{code}
\end{center}
\caption{\small{A typical codeword from a codebook of size $200$ (code
    7 in Figure \ref{fig:codes1}), corresponding to a
    tonic-subdominant chord progression.  The patch is composed of $2$
    bars and the pattern length was set to $8$ beats.  }}
\label{fig:code}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=.8\columnwidth]{encoding_per_beat}
\end{center}
\caption{\small{Encoding patterns of different sizes with a fixed size
codebook of 100 patterns. The size of the pattern is defined by the number of beats. 
Downbeat (bar alignment) information was not used for this experiment. % why not?
}}
\label{fig:perbeat}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=.99\columnwidth]{freqs}
\end{center}
\caption{\small{Usage proportions for all 200 codewords 
on the {\it artist20} test set (which comprises $71,832$
patterns).  Also shown are the usage proportions for the training set (``cowbell''), 
which are similar.
Note that even though all codewords are initialized from samples, some
are used only once in the training set, or not at all for test set.
}}
\label{fig:freqs}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=.9\columnwidth]{codes_lle}
\end{center}
\caption{\small{ Locally linear embedding (LLE) visualization of the
    codebook described in Section \ref{sec:visu}.  Each codeword is
    represented by a point on the scatter plot.  Randomly selected
    codewords from the embedded space are displayed over the plot.
    %FIXME: does this figure really provide any new insight?
  }}
\label{fig:lle}
\end{figure}

\begin{figure}[t]
\begin{center}
\includegraphics[width=.9\columnwidth]{code_variance}
\end{center}
\caption{\small{Average variance in time of a codebook. The vertical
axis cuts at the $63$rd pattern, a rough estimate of the number of 
codewords consisting of sustained chords.
}}
\label{fig:code_var}
\end{figure}

\begin{figure}[t]
\begin{center}
\includegraphics[width=.8\columnwidth]{codes_8bars}
\end{center}
\caption{\small{Sample of longer codewords spanning over $8$
    bars. Codewords selected randomly from the codewords used in the
    test set. Percentage of use display in parentheses.  Most patterns
    consist of sustained note or chords, but code 0 shows a repeating
    two chord pattern and code 4 contains two bars of a
    1$\rightarrow$1$\rightarrow$1$\rightarrow$2 progression.}}
\label{fig:codes8bars}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=.9\columnwidth]{close_patterns1}
\end{center}
\caption{\small{Cluster around centroid presented in
Figure \ref{fig:code}. Taken from the artist20 dataset, the cluster
size is actually $639$. Shown samples were randomly selected.
}}
\label{fig:cluster}
\end{figure}

\begin{figure}[t]
\begin{center}
\includegraphics[width=.8\columnwidth]{close_patterns_diff}
\end{center}
\caption{\small{First three patterns of figure \ref{fig:cluster}
($2$nd line) presented with the centroid from Figure \ref{fig:code}
($1$st line) and the absolute difference between both ($3$rd line).
}}
\label{fig:cluster_diff}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=.9\columnwidth]{song_encoded}
\end{center}
\caption{\small{Good Day Sunshine by \textit{The Beatles}.
Original song and its encoding using a $200$ entry codebook learning
$2$ bars at a time.
}}
\label{fig:encodesong}
\end{figure}

\section{Applications}\label{sec:exps2}
In this section we investigate the application of beat-chroma
codebooks to two tasks unrelated to clustering.
%
In the first experiment, we show that the codebooks retrain
information related to bar alignment and can therefore be used to
discover the bar boundaries for new songs.  The second experiment
shows that we can discriminate artists based on the encoding of their
songs, implying that some artists have a tendency to use different
beat-chroma patterns than others.

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{cc}
\toprule
Offset & \% of times chosen \\
\midrule
0 & $\mathbf{62.6}$\\
1 & $16.5$\\
2 & $09.4$\\
3 & $11.5$\\
\bottomrule
\end{tabular}
\end{center}
\caption{\small{
Results from the song segmentation experiment.
The correct offset is $0$.
Baseline performance by choosing the offset randomly gives an accuracy of $25$\%.
}}
\label{tab:offset}
\end{table}

@@

\begin{figure}[t]
%\begin{center}
\includegraphics[width=.9\columnwidth]{conf_mat_per_artist}
%\end{center}
\caption{\small{Confusion matrix for artist recognition task.}}
\label{fig:conf_mat}
\end{figure}

@@

% ignoring the two following figures
\iffalse
\begin{figure}[t]
\begin{center}
\includegraphics[width=.4\columnwidth]{metallica_pattern}
\end{center}
\caption{\small{Most typical pattern for Metallica.
COULD BE MIXED WITH TORI AMOS FIGURE.
}}
\label{fig:metallica}
\end{figure}

\begin{figure}[t]
\begin{center}
\includegraphics[width=.4\columnwidth]{toriamos_suzannevega_pattern}
\end{center}
\caption{\small{Most typical pattern for Tori Amos and Suzanne Vega.
    Can we show the top 2 or 4 for each to get a better idea?}}
\label{fig:amosvega}
\end{figure}
\fi

\begin{figure}[t]
  \centering
  \subfloat[][\small{Metallica}]{\label{fig:metallica2}\includegraphics[width=0.32\columnwidth]{metallica_pattern}}
  \hspace{5mm}                
  \subfloat[][\small{Tori Amos / Suzanne Vega}]{\label{fig:amosvega2}\includegraphics[width=0.32\columnwidth]{toriamos_suzannevega_pattern}}
  \vspace{2mm}
  \caption{\small{Typical patterns for different artists.}}
  \label{fig:typicalpat}
\end{figure}

\section{Conclusion and Future Work}
We presented a practical method to perform large-scale clustering of
harmonic patterns. We assessed the basic properties of the method through
experiments on a large collection of music, and 
demonstrated that these codebooks capture musically meaningful
patterns, at least to contemporary pop music.
%
 We suggested many ways
to interpret the data %(such as LLE)
and assert the merits of the codebook
through further experiments.
We discussed the possibility to move to even larger scales
and provided our source code.\footnote{Code not yet released to preserve
submission's anonymity}
%
Finally, we demonstrated the utility of unsupervised clustering to MIR
tasks that are typically tackled in a supervised setting.
